---
title: "Introduction to PscoresPlus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to PscoresPlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**PscoresPlus** is an R package for summarising and interpreting treatment hierarchies from
    network meta-analysis (NMA) using extended P-scores. It aims to help users explore how rankings change
    when different clinically important values (CIVs) are assumed, including uncertainty in the choice
    of CIV. The package quantifies hierarchy stability using P-score residuals and Precision of
    Treatment Hierarchy (POTH). It provides the Area Under the P-score Curve (AUPC) to rank treatments
    across a range of CIVs and multiple outcomes simultaneously. It also includes graphical tools to
    compare and visualize treatment hierarchies across outcomes and CIV ranges.

## Installation

```{r install, eval = FALSE}
# Install from CRAN (when available)
install.packages("PscoresPlus")

# Or install the development version from GitHub
# install.packages("remotes")
remotes::install_github("sofia-tsokani/PscoresPlus")
```

## Loading the Package

```{r load}
library(PscoresPlus)
```

## Datasets

The package includes two built-in datasets from a published network meta-analysis of antipsychotic drugs in schizophrenia (Leucht et al., 2013). Both have been included as `netmeta` objects and contain treatment effects (TE), standard errors (seTE), treatment comparisons, and study labels. Treatment effects are expressed as standardized mean differences (SMD).

> **Reference:** Leucht S, Cipriani A, Spineli L, et al. Comparative efficacy and tolerability of 15 antipsychotic drugs in schizophrenia: a
multiple-treatments meta-analysis. *Lancet*. 2013;382(9896):951–962.
https://doi.org/10.1016/S0140-6736(13)60733-3

### Efficacy (`efficacy`)

A network of **167 RCTs** comparing **15
antipsychotic treatments and placebo** on treatment efficacy (overall change in symptoms as measured with several scales). Although efficacy is inherently beneficial, it is treated as a **harmful outcome**:  a greater reduction in symptoms indicates better treatment performance. Treatment effects are expressed as standardized mean differences (SMD).

```{r}
#load the package
data("efficacy")
```

### Weight Gain (`weight_gain`)

A network of **86 RCTs** comparing **15
antipsychotic treatments and placebo**. This dataset
focuses on **weight gain** as a continuous outcome. Weight gain is treated as a **harmful outcome**, so treatments associated with smaller increases in weight are considered more favorable. Treatment effects are expressed as standardized mean differences (SMD).

```{r}
#import dataset
data("weight_gain")
```

## Single Outcome Example

When analysing a single outcome, supply the outcome of interest as a single-element list of the corresponding `netmeta` object, the CIV to be evaluated, and the outcome type ( `"H"` for harmful or `"B"` for beneficial). Set `excel = TRUE` to export the results directly to an Excel file.

The function provides:

- *Extended P-scores* per treatment
- *Treatment rankings* (1 = best)
- *Precision of Treatment Hierarchy (PoTH)*
- *P-score residuals*


### Efficacy Outcome

In this example, we compute extended P-scores for a single outcome (efficacy), treating it as a **harmful** outcome with a CIV of **0.5**.

```{r single-outcome}
out_eff <- pscore_civs(
  x           = list(efficacy),
  CIVs        = 0.5,
  correlation = NULL,
  type        = "H"
)

out_eff
```

### Weight Gain Outcome

Here we repeat the analysis for weight gain, a harmful outcome, using a CIV of **-0.5**.

```{r weight-gain}
data("weight_gain")

out_wg <- pscore_civs(
  x           = list(weight_gain),
  CIVs        = -0.5,
  type        = "H", 
  excel = TRUE #this allows to save results in excel file
)

out_wg
```

## Multiple Outcomes Example

When analysing multiple outcomes simultaneously, you must supply:

1. A *list of `netmeta` objects*, one per outcome
2. A *list of CIV vectors*, one per outcome
3. A *correlation matrix* between outcomes
4. A *`type` vector* specifying whether each outcome is harmful (`"H"`) or beneficial (`"B"`)

### Defining the Correlation Matrix

We assume a correlation of **−0.5** between efficacy and weight gain:

```{r cor-matrix}
cor_mat <- matrix(
  c( 1.0, -0.5,
    -0.5,  1.0),
  nrow   = 2,
  byrow  = TRUE
)
cor_mat
```

### Single CIV per Outcome

```{r multi-single-civ}
out1 <- pscore_civs(
  x           = list(efficacy, weight_gain),
  CIVs        = list(0.5, -0.5),
  correlation = cor_mat,
  type        = c("H", "H")
)

out1
```

### Multiple CIVs per Outcome

You can supply **multiple CIV values** for any outcome. The function evaluates all the CIV combinations. In the example below, weight gain has two CIVs (−0.3 and −0.5), yielding *2 combinations* in total.

```{r multi-multi-civ}
out2 <- pscore_civs(
  x           = list(efficacy, weight_gain),
  CIVs        = list(0.5, c(-0.3, -0.5)),
  correlation = cor_mat,
  type        = c("H", "H"),
  excel       = FALSE
)

out2
```

> Tip: Set the argument `excel = TRUE` to automatically export results to a timestamped Excel file (e.g., `pscore_civs_20260421_221800.xlsx`) in your working directory.

## Understanding the Output

The returned object is of class `"pscore_civs"` and includes the following items:

| Column | Description |
|---|---|
| `Treatment` | Treatment label |
| `Pscore` | Raw extended P-score under that CIV combination |
| `ranking` | Rank of the treatment (1 = best) |
| `residuals` | P-score residuals |
| `poth` | Precision of Treatment Hierarchy (PoTH) |


## Visualising P-scores Across CIVs

The `plot_pscores()` function plots extended P-scores or residual P-scores
across a range of CIV values for a single outcome. The top `top_n` treatments are highlighted in colour; a dashed reference line marks the mean P-score or approximately zero for residuals.
As with `pscore_civs()`, supply the outcome as a single-element list containing
the `netmeta` object, a numeric vector of CIV values to evaluate, and the
outcome type (`"H"` for harmful or `"B"` for beneficial).

### Extended P-scores

```{r, fig.width = 10, fig.height = 6, fig.align = "center", out.width = "100%"}
plot_pscores(
  x         = list(efficacy),
  CIVs      = seq(0, 0.5, 0.1),
  type      = "H",
  top_n     = 5,
  residuals = FALSE
)


```

### Residual P-scores

Residual P-scores centre each treatment's P-score by subtracting the row mean
across all treatments at each CIV value.

```{r, fig.width = 10, fig.height = 6, fig.align = "center", out.width = "100%"}
plot_pscores(
  x         = list(efficacy),
  CIVs      = seq(0, 0.5, 0.1),
  type      = "H",
  top_n     = 5,
  residuals = TRUE
)
```

## Area Under the P-score Curve (AUPC)

The `aupc()` function summarises treatment performance across an entire CIV
grid by computing the **Area Under the P-score Curve (AUPC)** for each
treatment. AUPC can be interpreted as the average probability that a treatment
beats all others over the specified range of CIV values.
The function takes a `pscore_civs` object as input; set `excel = TRUE` to export results to an Excel file.

### Single Outcome (Efficacy)

```{r}
#Calculating AUPC for outcome efficacy with a grid of 50 values between 0 and 0.5.
out_eff <- pscore_civs(
  x           = list(efficacy),
  CIVs        = list(seq(0, 0.5, length.out = 50)),
  correlation = NULL,
  type        = "H"
)

aupc(out_eff)
```

### Single Outcome (Weight Gain)

```{r}
#Calculating AUPC for outcome weight gain with a grid of 50 values between 0 and 0.5.
out_wg <- pscore_civs(
  x           = list(weight_gain),
  CIVs        = list(seq(-0.5, 0, length.out = 50)),
  correlation = NULL,
  type        = "H"
)

aupc(out_wg)
```

### Multiple Outcomes (Efficacy and Weight Gain)

```{r}
#correlation matrix between the two outcomes (the assumed correlation between efficacy and weight gain is −0.5
cor_mat <- matrix(c( 1.0, -0.5,
                    -0.5,  1.0),
                  nrow  = 2,
                  byrow = TRUE)

out <- pscore_civs(
  x           = list(efficacy, weight_gain),
  CIVs        = list(seq(0, 0.5, length.out = 50),
                     seq(-0.5, 0, length.out = 50)),
  correlation = cor_mat,
  type        = c("H", "H")
)

aupc(out)
#aupc(out, excel=TRUE) # Export results to a timestamped Excel file
```

## Comparing AUPC and P-score Rankings

The `plot_aupc_pscore()` function produces a scatter plot comparing treatment
rankings derived from **AUPC** and **extended P-scores** at a specific CIV. An
identity line (y = x) is drawn as a reference, and treatments are labelled
directly on the plot, making it easy to spot where the two ranking approaches
agree or diverge.

```{r, fig.width = 9, fig.height = 6, fig.align = "center", out.width = "100%"}
# Extended P-scores at a single CIV of 0.30
eff_rank <- pscore_civs(
  x           = list(efficacy),
  CIVs        = 0.30,
  correlation = NULL,
  type        = "H"
)

# AUPC over a CIV range of 0.20 to 0.40
pscore_range <- pscore_civs(
  x           = list(efficacy),
  CIVs        = list(seq(0.2, 0.4, 0.1)),
  correlation = NULL,
  type        = "H"
)
aupc_res <- aupc(pscore_range)

plot_aupc_pscore(
  aupc         = aupc_res,
  pscore       = eff_rank,
  outcome_name = "Efficacy"
)
```

## Heatplot of Treatment Rankings

The `pscores_heatplot()` function visualises treatment rankings across a grid
of CIV combinations for **exactly two outcomes**. Each facet represents one
treatment, with the two CIV axes on x and y and ranking shown as a colour
gradient. This complements the AUPC summary by revealing the specific CIV
regions where a treatment performs better or worse.

```{r, fig.width = 12, fig.height = 8, fig.align = "center", out.width = "100%"}
pscores_heatplot(
  out,
  title    = "Efficacy & Weight Gain",
  outcome1 = "Efficacy",
  outcome2 = "Weight Gain"
)
```

> **Note:** `pscores_heatplot()` requires a `pscore_civs` object with exactly
> two outcomes. The `out` object computed in the previous section is used here
> directly.


## POTH Bubble Plot

The `pscores_pothplot()` function visualises **Precision of Treatment Hierarchy
(POTH)** values across a grid of CIV combinations for exactly two outcomes as a
bubble plot. Bubble size reflects POTH magnitude, and the maximum and minimum
values are optionally highlighted in distinct colours, revealing the CIV regions
where treatment rankings are most or least precise.

The function returns a list with two elements: `$plot` (the `ggplot2` object)
and `$grid` (the displayed CIV combinations with their POTH values).

```{r, fig.width = 9, fig.height = 7, fig.align = "center", out.width = "100%"}
g <- pscores_pothplot(
  x        = out,
  title    = "POTH Plot",
  outcome1 = "Efficacy",
  outcome2 = "Weight Gain"
)

g$plot
```

For large grids, use `newgridsize` to downsample to an n × n quantile grid for
a cleaner visualisation:

```{r, fig.width = 9, fig.height = 7, fig.align = "center", out.width = "100%"}
g2 <- pscores_pothplot(
  x           = out,
  newgridsize = 10,
  title       = "POTH Plot",
  outcome1    = "Efficacy",
  outcome2    = "Weight Gain"
)

g2$plot
```
```{r session-info}
sessionInfo()
```
